---
title: "Parity and AD specific neurodegeneration"
author: "Clazz"
toc: true
number-sections: true
format: 
  docx:
    code-fold: true
editor: visual
bibliography: references.bib
csl: elsevier-vancouver.csl
---

```{r, include = F}
library(readxl)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("psych")
library(psych)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("ggpubr")
library(ggpubr)
#install.packages("broom")
library(broom)
#install.packages("ggeffects")
library(ggeffects)
#install.packages("MASS")
library(MASS)
#install.packages("mice")
library(mice)
#install.packages("VIM")
library(VIM)
#install.packages("car")
library(car)
#install.packages("sjPlot")
library(sjPlot)
library(ggrain)
#install.packages("emmeans")
library(emmeans)
#install.packages("skimr")
library(skimr)
library(glmnet)
#install.packages("lavaan")
library(lavaan)
#install.packages("patchwork")
library(patchwork)
#install.packages("lme4")
library(lme4)
#install.packages("lmerTest")
library(lmerTest)
library(gtsummary)
library(broom.mixed)
library(purrr)
library(huxtable)
```

```{r setup, include=FALSE}
# Register an inline hook:
options(scipen = 1, digits = 3)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 8,
  out.width = "100%",
  dpi = 300
)
```

```{r, include=FALSE}
## Plot aesthetics


call_aesthethics <- function(text_size){
  th <- theme(   panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank(),
                 axis.line = element_line(linewidth = 0.5),
                 legend.position = 'right',
                 legend.text = element_text(size= text_size, family="Helvetica"),
                 text = element_text(size= text_size, family="Helvetica"),
                 strip.text.x = element_text(size = rel(0.90)),
                 strip.text.y = element_text(size = rel(0.90)),
                 axis.title.x = element_text(vjust=-0.3),
                 plot.title = element_text(hjust = 0.5, vjust = 0),
                 axis.ticks = element_line(linewidth = 0.4),
                 axis.text.x.bottom  = element_text(size = rel(0.90), margin = unit(c(t = 2.5, r = 0, b = 0, l = 0), "mm")),
                 axis.title.y = element_text(vjust = 1),
                 axis.text.y = element_text(size = rel(0.90), margin = unit(c(t = 0, r = 2.5, b = 0, l = 0), "mm")),
                 axis.ticks.length = unit(-1.2, "mm"),
                 axis.text.x.top = element_text(size = rel(0.90), margin = unit(c(t = 0, r = 0, b = 2.5, l = 0), "mm")))
  return(th)
}


th <- call_aesthethics(13)

```

# Abstract

Background and objectives: Epidemiological studies exploring the effect of parity on Alzheimer’s Disease (AD) yield mixed results, and little is known about how parity may modify the impact of AD pathology on brain and cognitive aging. We examined the effect of parity and AD pathology on cognitive and hippocampal volume (HV) changes in cognitively unimpaired (CU) postmenopausal women.

Methods: We included 254 CU postmenopausal women from the ALFA+ study who completed two visits spaced by \~3 years (mean = 3.38, sd = 0.55) with reproductive data, cognitive testing (Preclinical Alzheimer Cognitive Composite), CSF biomarkers (Aβ42 and Aβ40), and structural MRI. We used linear mixed-effects models to investigate the interactive effects of parity, time, and Aβ status (based on CSF Aβ42/Aβ40) on cognition and HV. APOE-ε4 status, age, and total intracranial volume were included as covariates.

Results: A significant interaction was observed between parity and Aβ status on cognitive decline and HV. In women with a positive amyloid status, higher parity was associated with a steeper cognitive decline (p=0.033) and lower HV across timepoints (p = 0.036). Sensitivity analyses indicated that multiparous (2+ childbirths) women showed the strongest association between Aβ status and these outcome variables. Mediation analyses showed that in multiparous women, the effect of Aβ status on cognition is partly mediated by HV.

Discussion: In CU postmenopausal women, higher parity interacts with AD pathology to negatively influence cognitive changes and hippocampal volume. These findings suggest that parity impacts the resilience to AD pathology in preclinical AD stages.

# Introduction

Alzheimer’s disease (AD) is a neurodegenerative disorder characterized by the accumulation of pathophysiological abnormalities leading to neurodegeneration and progressive memory loss. β-amyloid (Aβ) protein accumulation, a hallmark neuropathological marker of AD, is thought to trigger a pathological cascade, leading to further accumulation of tau protein, neurodegeneration, and cognitive impairment @jack2013. This pathological process particularly affects the hippocampus early in the disease course, making hippocampal volume (HV) a hallmark predictor of cognitive decline in AD @apostolova2006. Being biologically female is one of the main non-modifiable risk factors for AD @riedel2016. Females who develop AD show a worse cognitive trajectory after mild cognitive impairment (MCI) diagnosis @holland2013, faster atrophy rate in the hippocampus @ardekani2016, and may benefit less from amyloid-antibody based therapies @andrews2025. These differences in pathology suggest that sex-dependent pathways or biological processes may drive distinct pathological trajectories. In particular, pregnancy is a female-specific event whose long-term effect on the brain remains poorly understood.

Pregnancy is a complex biological process which provokes significant changes to hippocampal structure and volume @puri2023. Animal models show decreased neurogenesis in the hippocampus during gestation and the postpartum period [@pawluski2007; @eid2019]. Whilst clinical evidence suggests that gestation negatively affects HV up to at least 6 years postpartum @martínez-garcía2021, whether this effect persists in older adults is unknown. These potentially long-lasting structural alterations may be particularly consequential in the context of early AD pathogenesis in which HV is also affected @cacciaglia2025. Therefore, parity-induced alterations to the hippocampus may create a neurobiological context that alters the extent to which AD pathology affects cognitive trajectories and hippocampal volume in later life.

The effect of parity on AD risk and cognitive decline remains poorly understood, and existing evidence paints a complex picture. Some epidemiological studies suggest a potential association between grand multiparity (i.e., having 4+ children) and increased risk of AD-related cognitive impairment @Colucci2006; @bae2020. However, others have shown no association @wolfova2024, the opposite association @fox2013, or a non-linear effect showing that, in addition to gran multiparity, nulliparity might also be associated with worse AD-related outcomes @keinert2023. Recent evidence indicates that parity modulates women’s cognitive performance differently depending on their disease status. Specifically, a recent cross-sectional study of 446 women (207 CU, 155 mild cognitive impairment, and 84 clinically diagnosed dementia) found that higher parity was associated with enhanced executive function but reduced episodic memory only among the women with dementia and not the cognitively healthy women @lee2025.

However, these studies did not include AD-specific biomarkers, meaning disease etiology was uncertain and prevented examination of how parity may influence cognitive and structural trajectories earlier in the AD trajectory. To our knowledge, only two studies have included AD biomarkers or neuropathological change as dependent variables, with mixed results. A post-mortem histopathological study of 42 women spanning the neuropathological spectrum from normal aging to definite AD found that multiparity was associated with a greater number of neuritic plaques @Beeri2009a. However, analysis using Aβ-PET data from 248 women who were either CU or had clinically diagnosed MCI found no association between Aβ burden and parity or multiparity @jung2020. These inconsistent findings and the lack of examining whether parity interacts with Aβ pathology to affect other brain or cognitive outcomes, highlight the need for a more nuanced examination of the interplay between parity and AD pathology in women in the AD preclinical stage.

To elucidate the aforementioned landscape of mixed results, we investigated the interplay between parity, Aβ status, cognition and HV in postmenopausal cognitively unimpaired women. We hypothesized that, rather than a direct association of parity to Aβ load, parity would interact with Aβ load in its association with HV and cognitive decline.

# Methods

## Study population

We included postmenopausal women who completed the baseline and first longitudinal visit (with an average time between visits of 3.32 years) from the ongoing ALFA+ study (Clinicaltrials.gov, NCT01835717). ALFA+ is a research cohort of middle-aged cognitively unimpaired participants, many of whom are offspring of AD patients, who have been deeply characterized by clinical interviews, lifestyle and risk factor questionnaires, cognitive testing, CSF biomarkers, and neuroimaging procedures, including magnetic resonance imaging (MRI), and Aβ positron emission tomography (PET). All of these procedures are repeated every 3 years with the main aim of identifying the earliest pathophysiological changes in the preclinical AD continuum @molinuevo2016. Briefly, ALFA+ inclusion criteria were: 1) participants who had previously participated in the 45-65/FPM2012 study (ALFA parent cohort); 2) age between 45 and 75 years at the moment of inclusion in the cohort; 3) long-term commitment to undergo all tests and study procedures (MRI, PET, and lumbar puncture). Exclusion criteria at baseline included: 1) cognitive impairment (Clinical Dementia Rating \[CDR\] \> 0, Mini-Mental State Examination \[MMSE\] \< 26, semantic fluency \< 12); 2) any significant systemic illness or unstable medical condition which could lead to difficulty complying with the protocol; 3) any contraindication to any test or procedure; 4) family history of monogenic AD. In addition, women with a history of cardiovascular-related gestational disease, such as preeclampsia, were excluded from our sample for analysis. Sex and menopausal status, defined as absence of menstruations for 12 consecutive months, were self disclosed. Women who reported being premenopausal or perimenopausal were not included in our analysis either.

The study was approved by the Independent Ethics Committee “Parc de Salut Mar”, Barcelona, and all participants gave written informed consent. Baseline visits took place between 2016 and 2019, and the follow-up visit between 2019 and 2022.

The sample consisted of 254 participants at baseline visit, and 211 at follow-up (Table 1). Out of the 43 participants who dropped out of the study between visits, 30 withdrew for personal reasons, 3 became unresponsive, 2 died, 5 showed incidental l findings on their structural MRI, and 3 were excluded due to mild cognitive impairment clinical diagnosis.

## Reproductive history variables

Parity was defined as the number of biological children and used as a continuous variable in the main analyses and as a categorical variable in sensitivity analyses. When categorical, parous groups were defined as follows: nulliparous (0 childbirths), primiparous (1 childbirth), and multiparous (2+ childbirths). This information was self-reported by individuals during an interview with a clinician at the baseline visit.

## Cognitive measures

A modified Preclinical Alzheimer’s cognitive composite (mPACC) was the main cognitive outcome of this study and was computed, including the Total Paired Recall (TPR) and Total Delayed Free Recall scores of the Memory Binding Test @buschke2017, the Coding subtest of the Wechsler Adult Intelligence Scale-Fourth Edition (WAIS-IV), and semantic fluency, as defined in previous works @lópez-martos2025. Z-scores were computed for Visit 1 and 2 and used in the linear regressions. When using score difference between visit 1 and 2 for the mediation analysis, change was computed by subtracting baseline scores from follow-up.

## AD biomarkers

CSF collection, processing, and storage in the ALFA+ study have been described previously @Milà-Alomà2020. Briefly, CSF Aβ42 and Aβ40 were measured with the exploratory Roche NeuroToolKit immunoassays (Roche Diagnostics International Ltd, Rotkreuz, Switzerland) on a cobas e 601 module. Measurements were performed at the Clinical Neurochemistry Laboratory, Sahlgrenska University Hospital, Molndal, Sweden. Aß status (Aß+, Aß–) was defined using the cutoff of 0.071 for the ratio Aß42/40 @Milà-Alomà2020, where any participant below this cutoff was considered to be Aβ+.

## Imaging data acquisition and preprocessing

Automatic Segmentation of Hippocampal Subfields (ASHS) software @Yushkevich2015 was used on the T1, T2 and inversion recovery (IR) weighted images to segment the hippocampal formation in the following sub-regions: Brodmann areas 35 and 36, cornu Ammonis (CA) 1, 2 and 3, dentate gyrus (DG), entorhinal cortex (ERC), parahippocampal cortex (PHC), subiculum (SUB), and the hippocampal sulcus (SUL). All segmentations were visually inspected before proceeding with the statistical analyses.

Hippocampal values are reported in cm3 and were obtained at both baseline and follow-up visits.

Total Intracranial volume (TIV) was estimated using deformation fields obtained during the nonlinear registration of each subject’s T1-weighted MRI to a whole-brain template. The FSL Brain Extraction Tool @smith2002 was applied to the template to create an intracranial mask, which was warped into each subject’s MRI at native space to obtain the TIV measurement.

## Statistical analyses

Using baseline data, we used linear regression models to examine the association between parity (exposure) and Aβ status (outcome).

We then used linear mixed-effects models to assess the joint effects of parity and Aβ status (exposures), on HV and cognition (outcomes) over both visits, and test if this association is dependent on time. These models included a three-way interaction term among parity, Aβ positivity, and time (the interval between Visit 1 and Visit 2 in years) and all lower-order interactions. Cognitive analyses were adjusted for APOEe4 carriership, years of education, and age at baseline visit. HV analyses were additionally adjusted for scaled TIV.

Analyses that showed a significant parity x Aβ status interaction were reproduced using parity as a categorical variable to examine if the effect was driven by parity itself or if each additional pregnancy seemed to have a cumulative effect. For all groups we conducted post-hoc comparisons using estimated marginal means. Specifically, pairwise comparisons between parity groups were examined separately within each Aβ status category, with p-values adjusted for multiple comparisons using the Tukey method.

Additional linear regressions were also performed to explore the interaction between parity and Aβ positivity on individual hippocampal subfield volumes using baseline data. Given the exploratory nature of these analyses, results are reported without correction for multiple comparisons.

Statistical significance was set at p \< 0.05. Analyses were conducted using R software, version 4.5.1 (The R Foundation for Statistical Computing, Vienna, Austria). The package ‘emmeans’ was used to calculate estimated marginal mean for direct group comparison of predicted effects @lenth2017. The package MICE was used for missing data imputation @buuren2011. The package tidiverse was used for data processing and plotting @wickham2019. The package gtsummary was used for formatting tables @sjoberg2021. The package ggeffects was used to extract predicted values from regression models @lüdecke2018. The package lme4 was used for performing mixed effect linear regression models @bates2015.

Rates of missing data in exposures (parity, Aβ) or outcome variables (mPACC, HV) at each time point are reported (Table 1).

To elucidate the mechanisms underlying the parity-cognition relationship, we performed mediation analyses testing whether HV mediates the association between Aβ status and mPACC based on parity, given the well-established link between hippocampal integrity and cognitive function in preclinical and prodromal AD. We proceeded with the mediation model in whole group, then multiparous women only, which consisted of three pathways: (1) the effect of Aβ status on baseline HV (path a), (2) the effect of baseline HV on mPACC difference while controlling for Aβ status (path b), and (3) the direct effect of Aβ status on mPACC difference after accounting for the mediator (path c'). The indirect effect was calculated as the product of paths a and b. The package ‘mediation’ @tingley2014 was used for mediation analysis.

Statistical significance of the indirect effect was assessed using bootstrap confidence intervals with 5000 iterations, and the seed was set to 123. Covariates included age at baseline, APOE-ε4 carrier status, education, time between cognitive tests, and scaled TIV. HV was also scaled in this case due to being on a vastly different scale than the rest of the covariates.

# Results

```{r, include=FALSE}
# creating basic dataset

data_naked<-read_delim("G:/My Drive/Data/ALFA+/data_cleaned/ALFA_dataset_basic.csv", 
    delim = ",", escape_double = FALSE, trim_ws = TRUE)%>%
  mutate(sex=factor(sex),
         APOE_binary = factor(APOE_binary), 
         subject=as.numeric(subject),
         parous_group= factor(parous_group))

alfa_subjects<-data_naked%>%
  dplyr::select(subject, sex)

#adding scan dates for descriptives, horizontal 
date_mri_v1<-read_excel("G:/My Drive/Data/ALFA+/data_raw/neuroimaging/scan_dates/bx_scandates_ALFA_PLUS_V1_20231120_20250210_135308.xlsx")%>%
  dplyr::select(subject_label, scandate)%>%
  mutate(subject_label = as.numeric(subject_label))%>%
  rename(subject = subject_label,
         scandate_v1 = scandate)

date_mri_v2<-read_excel("G:/My Drive/Data/ALFA+/data_raw/neuroimaging/scan_dates/bx_scandates_ALFA_PLUS_V2_20230518_20250210_135351.xlsx")%>%
  dplyr::select(subject_label, scandate)%>%
  mutate(subject_label = as.numeric(subject_label))%>%
  rename(subject = subject_label,
         scandate_v2 = scandate)

date_mri_long<-full_join(date_mri_v1, date_mri_v2)%>%
  mutate(scandate_v1 = as.Date(scandate_v1, format = "%Y-%m-%d"),
         scandate_v2 = as.Date(scandate_v2, format = "%Y-%m-%d"),
         diff_mri = as.numeric(difftime(scandate_v2, scandate_v1, units = c("days")))/365.25)

#adding MRI sessions dates to the dataset
data<-left_join(data_naked, date_mri_long)%>%
  mutate(DOB = as.Date (DOB, format = "%Y-%m-%d"),
         age_v1_mri = as.numeric(difftime(scandate_v1, DOB, units= c("days")))/365.25,
         age_v2_mri = as.numeric(difftime(scandate_v2, DOB, units= c("days")))/365.25)

data_biomarkers<-read_delim("G:/My Drive/Data/ALFA+/data_cleaned/biomarkers_CSF_processed.csv", 
    delim = ",", escape_double = FALSE, trim_ws = TRUE)%>%
  filter(Visit == "V1")%>%
  dplyr::select(-Visit)%>%
  mutate(A=factor(A))

data<-left_join(data, data_biomarkers)



```

```{r hippocampus_data, include = F}
ashs1<-read_excel("G:/My Drive/Data/ALFA+/data_raw/neuroimaging/ASHS/bx_ashs_volumes_ALFA_PLUS_V1_20231120_20250702_164509.xlsx")%>%
  dplyr::select(subject, region, side, volume)%>%
  group_by(subject, region) %>%
  summarize(volume = sum(volume), .groups = "drop")%>%
  pivot_wider(names_from = region, values_from = volume, names_prefix = "vol_")%>%
  mutate(Visit = 1,
         subject= as.numeric(subject))

ashs1b<-read_excel("G:/My Drive/Data/ALFA+/data_raw/neuroimaging/ASHS/bx_ashs_volumes_ALFA_PLUS_EXP_20250918_161906.xlsx")%>%
  dplyr::select(subject, region, side, volume)%>%
  group_by(subject, region) %>%
  summarize(volume = sum(volume), .groups = "drop")%>%
  pivot_wider(names_from = region, values_from = volume, names_prefix = "vol_")%>%
  mutate(Visit = 1,
         subject= as.numeric(subject))%>%
  filter(subject != 11331,
         subject != 10634,
         subject!= 13368)
  

ashs2<-read_excel("G:/My Drive/Data/ALFA+/data_raw/neuroimaging/ASHS/bx_ashs_volumes_ALFA_PLUS_V2_20230518_20250702_165801.xlsx")%>%
  dplyr::select(subject, region, side, volume)%>%
  group_by(subject, region) %>%
  summarize(volume = sum(volume), .groups = "drop")%>%
  pivot_wider(names_from = region, values_from = volume, names_prefix = "vol_")%>%
  mutate(Visit = 2,
         subject= as.numeric(subject))

data_ashs<-ashs1%>%
  bind_rows(ashs1b)%>%
  bind_rows(ashs2)%>%
  mutate(subject = as.numeric(subject),
         vol_HIPP = vol_CA1+vol_CA2+vol_CA3+vol_DG+vol_SUB)%>%
  right_join(., data)%>%
  mutate(diff_mri = ifelse(Visit==1, 0, diff_mri),
         vol_HIPP_cm = vol_HIPP/1000,
         vol_tiv_cm= vol_tiv/1000)
 
```

```{r cognition_data, include=F}
cognition<-read_excel("G:/My Drive/Data/ALFA+/data_raw/Cognition/PACCv1_to_v3(incomplete).xlsx")%>%
  dplyr::select(IdParticipante, Visit, Z_PACC, Z_FCSRTI_RT, Z_aWMS_DR_RS, Z_Wclave_totalscr, Z_FS_totalcnswr, FechaResultado)%>%
  rename(subject = IdParticipante)

time_diff<-read_excel("G:/My Drive/Data/ALFA+/data_raw/Cognition/PACCv1_to_v3(incomplete).xlsx")%>%
  dplyr::select(IdParticipante, Visit,FechaResultado)%>%
  rename(subject = IdParticipante)%>%
  filter(Visit != 3)%>%
  pivot_wider(names_from = Visit, values_from = FechaResultado, names_prefix = c("Fecha_Resultado"))%>%
  mutate(time_diff = as.numeric(difftime(Fecha_Resultado2, Fecha_Resultado1, units= c("days"))))%>%
  dplyr::select(subject, time_diff)

data_cog_long<- left_join(data, cognition)%>%
  mutate(age_cog = as.numeric(difftime(FechaResultado, DOB, units= c("days")))/365.25)%>%
  group_by(subject)%>%
  filter(Visit != 3) %>%
  filter(n()>=2)%>%
  relocate(Visit, .after=sex)%>%
  ungroup()%>%
  left_join(., time_diff)%>%
  mutate(time_diff = ifelse(Visit == 1, 0, time_diff),
         time_diff_years = time_diff/365.25)
```

```{r intext_descriptives, include=F}
disc_participants_alfa<- cognition%>%
  group_by(subject)%>%
  filter(n()==1, Visit==1)%>%
  right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  drop_na(Visit)

participants_v1_cog<-read_excel("G:/My Drive/Data/ALFA+/data_raw/Cognition/PACCv1_to_v3(incomplete).xlsx")%>%
  dplyr::select(IdParticipante, Visit, Z_PACC)%>%
  rename(subject = IdParticipante)%>%
  filter(Visit == 1)%>%
   right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  drop_na(Z_PACC)%>%
  nrow()

participants_v2_cog<-read_excel("G:/My Drive/Data/ALFA+/data_raw/Cognition/PACCv1_to_v3(incomplete).xlsx")%>%
  dplyr::select(IdParticipante, Visit, Z_PACC)%>%
  rename(subject = IdParticipante)%>%
  filter(Visit == 2)%>%
   right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  drop_na(Z_PACC)%>%
  nrow()
  
participants_v1_mri<-data%>%
  drop_na(scandate_v1)%>%
  filter(sex==2)%>%
  nrow

participants_v2_mri<-data%>%
  drop_na(scandate_v2)%>%
    filter(sex==2)%>%
  nrow


```

## Descriptive results

```{r, include = F}
alfa_participants_v1<-read_delim("G:/My Drive/Data/ALFA+/data_raw/Alfa Plus Long 2025__AlfaPlus_general_20250304095809(in).csv", delim= ";", col_names = TRUE)%>%
  rename(subject = IdParticipante)%>%
  right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  dplyr::select(subject,Age_V1)%>%
  drop_na()%>%
  nrow()

alfa_participants_v2<-read_delim("G:/My Drive/Data/ALFA+/data_raw/Alfa Plus Long 2025__AlfaPlus_general_20250304095809(in).csv", delim= ";", col_names = TRUE)%>%
  rename(subject = IdParticipante)%>%
  right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  dplyr::select(subject,Age_V2)%>%
  drop_na()%>%
  nrow()

alfa_subjects_dropout<-read_delim("G:/My Drive/Data/ALFA+/data_raw/Alfa Plus Long 2025__AlfaPlus_general_20250304095809(in).csv", delim= ";", col_names = TRUE)%>%
rename(subject = IdParticipante)%>%
  right_join(., alfa_subjects)%>%
  filter(sex==2)%>%
  filter(is.na(Age_V2))%>%
  dplyr::select(subject)

dropout<-read_delim("G:/My Drive/Data/ALFA+/data_raw/Alfa Plus Long 2025_2010_STF_20250304095809(in).csv", delim= ";", skip=1, col_names = TRUE, locale = locale(encoding = "Latin1") )%>%
  rename(subject = IdParticipante)%>%
  right_join(., alfa_subjects_dropout, join_by(subject))%>%
  dplyr::select(subject, STF_state, STF_reas, STF_reas5, STF_EspeIncumCri)%>%
  mutate(STF_reas = factor(STF_reas, levels = c("1", "2", "3", "4", "5", "6", "7"), labels = c("Death", "No response", "Participant withdrawal", "Clinician decision", "Promotor decision", "Protocol issue", "Other")),
         STF_state = factor(STF_state, levels = c("1", "2", "3", "4"), labels = c("Completed", "Failed preescreening", "Failed screening", "Discontinued")))

cognition1<-cognition%>%
  filter(Visit==1)%>%
  dplyr::select(subject, Z_PACC)

cognition2<-cognition%>%
  filter(Visit==2)%>%
  dplyr::select(subject, Z_PACC)

data_table<-data%>%
  filter(sex==2)%>%
  left_join(., ashs1)%>%
  mutate(vol_HIPP_v1 = vol_CA1+vol_CA2+vol_CA3+vol_DG+vol_SUB)%>%
  left_join(., cognition1)%>%
  rename(Z_PACC_v1 = Z_PACC)%>%
  dplyr::select(age_v1, kids_total, A, APOE_binary, education, vol_tiv, vol_HIPP_v1, Z_PACC_v1, subject)%>%
  mutate(kids_total= ifelse(kids_total==4, "4+", as.character(kids_total)))%>%
  left_join(., ashs2, join_by(subject))%>%
  mutate(vol_HIPP_v2 = vol_CA1+vol_CA2+vol_CA3+vol_DG+vol_SUB)%>%
 dplyr::select(age_v1, kids_total, A, APOE_binary, education, vol_tiv.x, vol_HIPP_v1, vol_HIPP_v2, Z_PACC_v1, subject)%>%
  left_join(.,cognition2)%>%
  rename(Z_PACC_v2 = Z_PACC)%>%
   dplyr::select(age_v1, kids_total, education,  A, APOE_binary, Z_PACC_v1, Z_PACC_v2, vol_HIPP_v1, vol_HIPP_v2, vol_tiv.x)%>%
  mutate(vol_tiv.x=vol_tiv.x/1000,
         vol_HIPP_v1 = vol_HIPP_v1/1000,
         vol_HIPP_v2 = vol_HIPP_v2/1000)
 

desc_table<-tbl_summary(data_table,
                        statistic = list(all_continuous() ~ "{mean} ({sd})"),
            by = kids_total, 
            missing = "no",
            label= list(kids_total = "Number of childbirths",
                     A= "Aβ positive, Count (%)",
                     APOE_binary = "APOE-E4 carriers, Count (%)",
                     age_v1 = "Age at first visit, Mean (SD)",
                     vol_tiv.x = "Baseline Total Intracranial Volume (cm3), Mean (SD)",
                     education = "Education (years), Mean (SD)",
                     vol_HIPP_v1 = "Baseline Hippocampal volume (cm3), Mean (SD)",
                     vol_HIPP_v2 = "Follow-up Hippocampal volume (cm3), Mean (SD)",
                     Z_PACC_v1 = "Baseline PACC z-score, Mean (SD)",
                     Z_PACC_v2 = "Follow-up PACC z-score, Mean (SD)"),
            
            value = list(APOE_binary= "1",
                         A= "1"))%>%
  add_n()%>%
  add_overall()%>%
  add_p()%>%
  modify_header(label = "**Variables**") %>% # update the column header
  bold_labels()



```

```{r, echo = F, warning=F}
desc_table
```

As shown in Table 1, 15.3% of our participants had no previous childbirths, most participants had 2 (49.6%) and only 3% had 4 or more. Participants with higher parity were significantly older . No differences were observed among groups for Aβ status, APOE-ε4 carrier status, HV, or mPACC performance.


As shown in Table 1, 15.3% of our participants had no previous childbirths, most participants had 2 (49.6%) and only 3% had 4 or more. Participants with higher parity were significantly older. No differences were observed among groups for Aβ status, APOE-ε4 carrier status, HV, or mPACC performance. 

```{r, include=F}
data_ashs_cog_long<-data_ashs%>%
  mutate(Visit=as.character(Visit))%>%
  left_join(.,cognition, join_by(subject, Visit,))

fit_cog_ashs<-lmer(Z_PACC~ vol_HIPP_cm*diff_mri + kids_total+ age_v1 + education + APOE_binary + vol_tiv_cm +(1|subject), data = data_ashs_cog_long[data_ashs_cog_long$sex==2,])
summary(fit_cog_ashs)
plot_model(fit_cog_ashs, type="int")
```


## Interplay between parity, Aβ status and cognition

```{r, include=F}

data_bio<-left_join(data, data_biomarkers)

fit_amyloid<-glm(A ~ kids_total +age_v1 + APOE_binary + education , data_bio[data_bio$sex==2 ,], family=binomial(link = "logit"))
summary(fit_amyloid)


```

Using baseline data, parity was not associated with Aβ status (p=`r tidy(fit_amyloid)$p.value[2]`) (eTable 1).

```{r, include=F}

fit_cog_PACC<-lmer(Z_PACC ~ kids_total*A*time_diff_years +age_v1 + APOE_binary + education +(1 | subject), data_cog_long[data_cog_long$sex==2,])
summary(fit_cog_PACC)


fit_cog_PACC_group<-lmer(Z_PACC ~ parous_group*A*time_diff_years +age_v1 + APOE_binary + education +(1 | subject), data_cog_long[data_cog_long$sex==2,])
summary(fit_cog_PACC_group)

trend_cog<-emtrends(fit_cog_PACC_group, ~ A * parous_group, var = "time_diff_years")
pairs(trend_cog, by= "parous_group", adjust = "tukey")
```

Using data across both visits (eTable 2), parity did not significantly relate to mPACC changes (p=`r tidy(fit_cog_PACC)$p.value[2]`).The interaction between parity and Aβ status on mPACC was not statistically significant (p= `r tidy(fit_cog_PACC)$p.value[10]`) The 3-way interaction between parity, time, and Aβ positivity revealed a significant effect on mPACC z-score change (β = `r tidy(fit_cog_PACC)$estimate[11]`, SE = `r tidy(fit_cog_PACC)$std.error[11]`, p=`r tidy(fit_cog_PACC)$p.value[11]`). Results suggest that Aβ- women showed less cognitive decline with more childbirths, whereas Aβ+ women showed steeper decline with more childbirths (Figure 1).

```{r, include=F}
library(ggeffects)

# Create grid of values for prediction - specify multiple time points
time_range <- range(data_cog_long$time_diff_years, na.rm = TRUE)
time_seq <- seq(from = time_range[1], to = time_range[2], length.out = 10)

emm_grid <- emmeans(fit_cog_PACC, ~ time_diff_years * kids_total * A,
                    at = list(time = time_seq),
                    data= data_cog_long)
emm_df <- as.data.frame(emm_grid)

slopes <- emtrends(fit_cog_PACC, ~ kids_total * A, var = "time_diff_years", data= data_cog_long, at = list(kids_total = c(0, 1, 2, 3, 4)))
slopes_df <- as.data.frame(slopes)

cog_p1<- ggplot(slopes_df, aes(x = kids_total, y = time_diff_years.trend, 
                            fill = as.factor(A))) +
  geom_col(position = "dodge", alpha = 0.8, color = "black", size = 0.3) +
  geom_errorbar(aes(ymin = time_diff_years.trend - SE, ymax = time_diff_years.trend + SE),
                position = position_dodge(width = 0.9), width = 0.3)+
  th+
  labs(x = "Number of childbirths", 
       y = "mPACC slope", 
       color = "Aβ Status",
       fill = "Aβ Status")+
  scale_fill_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Aβ-", "Aβ+"))+
  geom_text(data = data.frame(A = factor("Aβ+", 
                                                     levels = c("Aβ-", "Aβ+")), 
                              time_diff_years = Inf, 
                              Z_PACC = Inf, 
                              label = "p= 0.033*"),
            aes(x = time_diff_years, y = Z_PACC, label = label),
            size = 4, vjust = 1.5, hjust = 1.2, inherit.aes = FALSE)
```

```{r cog_plot_emmeans, echo=F, warning=F}
preds_pacc_group <- ggpredict(fit_cog_PACC_group, terms = c("time_diff_years", "parous_group", "A"))%>%
  as.data.frame()

preds_pacc_groupA0<-as.data.frame(preds_pacc_group)%>%
  rename(parous_group = group,
         A=facet)%>%
  filter(A==0)%>%
  mutate(A = "Aβ-",
         parous_group= factor(parous_group, levels = c("0", "1", "2+"), labels = c("Nulliparous", "Primiparous", "Multiparous")))


preds_pacc_groupA1<-as.data.frame(preds_pacc_group)%>%
  rename(parous_group = group,
         A=facet)%>%
  filter(A==1)%>%
  mutate(A = "Aβ+",
         parous_group= factor(parous_group, levels = c("0", "1", "2+"), labels = c("Nulliparous", "Primiparous", "Multiparous")))

cog_p2<-data_cog_long %>%
  filter(sex == 2, !is.na(A)) %>%
  mutate(A = factor(A, levels = c("0", "1"), labels = c("Aβ-", "Aβ+")),
         parous_group= factor(parous_group, levels = c("0", "1", "2+"), labels = c("Nulliparous", "Primiparous", "Multiparous"))) %>%
  ggplot(aes(x = time_diff_years, y = Z_PACC, color = A, group=subject)) +
  geom_line(alpha=0.2)+
  geom_line(data =preds_pacc_groupA0, aes(x=x, 
                                y=predicted, 
                                group=parous_group), size=1.3) +
  geom_line(data =preds_pacc_groupA1, aes(x=x, 
                                y=predicted, 
                                group=parous_group), size=1.3) +
  geom_ribbon(data =preds_pacc_groupA0, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=parous_group), inherit.aes = F, alpha=.09)+
  geom_ribbon(data =preds_pacc_groupA1, aes(x=x,  
                                  ymin= conf.low, 
                                  ymax = conf.high, 
                                  group=parous_group), inherit.aes = F, alpha=.09)+
  facet_wrap(~parous_group) +
  th+
  labs(x = "Time difference (years)", 
       y = "mPACC Z-score", 
       color = "Aβ status")+
  scale_color_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Aβ-", "Aβ+"))+
  geom_text(data = data.frame(parous_group = factor("Multiparous", 
                                                     levels = c("Nulliparous", "Primiparous", "Multiparous")), 
                              time_diff_years = Inf, 
                              Z_PACC = Inf, 
                              label = "p= 0.009*"),
            aes(x = time_diff_years, y = Z_PACC, label = label),
            size = 4, vjust = 1.5, hjust = 1.2, inherit.aes = FALSE)+
  geom_text(data = data.frame(parous_group = factor("Nulliparous", 
                                                     levels = c("Nulliparous", "Primiparous", "Multiparous")), 
                              time_diff_years = Inf, 
                              Z_PACC = Inf, 
                              label = "p= 0.514"),
            aes(x = time_diff_years, y = Z_PACC, label = label),
            size = 4, vjust = 1.5, hjust = 1.2, inherit.aes = FALSE)+
  geom_text(data = data.frame(parous_group = factor("Primiparous", 
                                                     levels = c("Nulliparous", "Primiparous", "Multiparous")), 
                              time_diff_years = Inf, 
                              Z_PACC = Inf, 
                              label = "p= 0.344"),
            aes(x = time_diff_years, y = Z_PACC, label = label),
            size = 4, vjust = 1.5, hjust = 1.2, inherit.aes = FALSE)
  
```

```{r cog_plot_all, echo=F, warning=F}

cog_p1/cog_p2+
  plot_annotation(tag_levels = 'A')

```

Reproducing the analysis across categorical parity groups (nulliparous, primiparous, multiparous) did not yield a significant 3-way estimate (eTable 3). However, group contrast indicated that the difference between Aβ- and Aβ+ participants' predicted slope was significant only in multiparous women (p=0.009) (Figure 1).

## Interplay between parity, Aβ status and hippocampal volume

```{r, include = F}

mod_ashs<-lmer(vol_HIPP_cm ~  kids_total*diff_mri*A +age_v1  + vol_tiv_cm + education + APOE_binary +(1|subject), data = data_ashs[data_ashs$sex==2 ,])
summary(mod_ashs)

mod_ashs_group<-lmer(vol_HIPP_cm ~  parous_group*diff_mri*A +age_v1  + vol_tiv_cm + education + APOE_binary +(1|subject), data = data_ashs[data_ashs$sex==2 ,])
summary(mod_ashs_group)


emm_parous_A <- emmeans(mod_ashs_group, 
                        specs = ~ parous_group * A,
                        data = data_ashs[data_ashs$sex==2,])
pairs(emm_parous_A, by= "parous_group")
```

No significant association between parity and HV was found (p= `r tidy(mod_ashs)$p.value[2]`) (eTable 4). However, the interaction between parity and Aβ status was significant (β = `r tidy(mod_ashs)$estimate[10]`, SE = `r tidy(mod_ashs)$std.error[10]`, p = `r tidy(mod_ashs)$p.value[10]`), showing that higher parity in Aβ- women was associated with a lower volume (Figure 2). Further exploration of this interaction in the hippocampal subfields can be found in the supplementary materials (eFigure 1).

The 3-way interaction between time, parity, and Aβ status was not significant (p = `r tidy(mod_ashs)$p.value[12]`).

When using categorical parous groups as predictors instead of continuous parity, the resulting model did not yield significant group estimates (eTable 5). However, group-contrast of estimated means indicated that the difference in predicted HV between Aβ- and Aβ+ participants was significant only in multiparous women (β=236, SE=81.8, 0.004) (Figure 2).

```{r hippo_plot, echo=F, warning=F}
cyan_gradient <- c("#1CA8B7", "#1CA8B7", "#1CA8B7", "#1CA8B7", "#1CA8B7")
orange_gradient <- c("#FF7043", "#FF7043", "#FF7043", "#FF7043", "#FF7043")

# Create a named vector that maps kids_total values to colors for each A group
color_mapping <- c(
  "Aβ-.0" = cyan_gradient[1], "Aβ-.1" = cyan_gradient[2], "Aβ-.2" = cyan_gradient[3], 
  "Aβ-.3" = cyan_gradient[4], "Aβ-.4+" = cyan_gradient[5],
  "Aβ+.0" = orange_gradient[1], "Aβ+.1" = orange_gradient[2], "Aβ+.2" = orange_gradient[3], 
  "Aβ+.3" = orange_gradient[4], "Aβ+.4+" = orange_gradient[5]
)
quantiles_95 <- function(x) {
  r <- quantile(x, probs=c(0.05, 0.25, 0.5, 0.75, 0.95))
  names(r) <- c("ymin", "lower", "middle", "upper", "ymax")
  r
}


hippo_p1<-data_ashs %>%
  filter(sex==2, Visit == 1) %>%
  drop_na(A) %>%
  mutate(A = factor(A, levels = c("0", "1"), labels = c("Aβ-", "Aβ+")),
         kids_total = ifelse(kids_total==4, "4+", as.character(kids_total)),
         combined_group = paste0(A, ".", kids_total)) %>%
  ggplot(aes(x = kids_total, y = vol_HIPP_cm, 
             group = combined_group, 
             colour = combined_group, 
             fill = combined_group)) +
  stat_summary(fun.data = quantiles_95, geom="boxplot", alpha=0.5, position=position_dodge(width = 0.9) ,width= 0.8, lwd=0.9)+
  geom_point(alpha = 0.5, position = position_dodge(width = 0.9), size = 2) +
  th +
  labs(x = "Number of childbirths", 
       y = "Total hippocampal volume (cm3)", 
       color = "Number of childbirths",
       fill = "Number of childbirths") +
  scale_fill_manual(values = color_mapping,
                    labels = c("0", "1", "2", "3", "4+", "0", "1", "2", "3", "4+")) +
  scale_color_manual(values = color_mapping,
                     labels = c("0", "1", "2", "3", "4+", "0", "1", "2", "3", "4+"))+
  theme(legend.position = "none")+
  annotate("text", x = Inf, y = Inf, label = "p = 0.036*", 
           hjust = 1.1, vjust = 1.5, size = 4)
  
```

```{r hippo_plot_emmeans, echo=F, warning=F}
preds_hip_lm <- ggpredict(mod_ashs_group, terms = c("parous_group", "A"))%>%
  as.data.frame()

hippo_p2<-
  
  preds_hip_lm%>%
  mutate(x=factor(x, levels = c("0", "1", "2+"), labels = c("Nulliparous", "Primiparous", "Multiparous")),
         group= factor(group, levels = c("0", "1"), labels= c("Aβ-", "Aβ+")))%>%
  ggplot(aes(x= x, y=predicted, group=group, colour = group))+
  geom_point(position = position_dodge(width = 0.8), size=3)+
  geom_errorbar(aes(x=x, ymin= conf.low, ymax = conf.high, group=group), position = position_dodge(width = 0.8), size=1.5, width=0.5)+
  th+
  labs(x = "Parity group", 
       y = "Total hippocampal volume (cm3)", 
       color = "Aβ status")+
  scale_color_manual(values = c("#1CA8B7", "#FF7043"),
                     labels = c("Aβ-", "Aβ+"))+
  geom_signif(
    annotation = "0.004**",  
    y_position = max(preds_hip_lm$conf.high) ,
    xmin = 2.8,
    xmax = 3.2,
    tip_length = 0.02,
    color = "black",
    size= 1.2,
    textsize = 4
  )
```

```{r hippo_plot_all, echo=F, warning=F}

hippo_p1/hippo_p2+
  plot_annotation(tag_levels = 'A')
```

## Post Hoc analysis

```{r, include=F}
PACC_wide<-cognition%>%
  dplyr::select(subject, Visit, Z_PACC, FechaResultado)%>%
  pivot_wider(values_from = c(Z_PACC, FechaResultado), names_from = Visit, names_prefix = "V")%>%
  mutate(Z_PACC_diff = Z_PACC_V2-Z_PACC_V1,
         FechaResultado_V1= as.Date(FechaResultado_V1, format = "%Y-%m-%d"),
         FechaResultado_V2= as.Date(FechaResultado_V2, format = "%Y-%m-%d"),
         cog_diff= as.numeric(FechaResultado_V2-FechaResultado_V1)/365.25)%>%
  dplyr::select(subject, Z_PACC_diff, cog_diff, Z_PACC_V2, Z_PACC_V1)


data_all<-data%>%
  filter(sex==2)%>%
  left_join(.,PACC_wide)%>%
  left_join(., ashs1)%>%
  mutate(vol_HIPP = vol_CA1+vol_CA2+vol_CA3+vol_DG+vol_SUB)%>%
  mutate(kids_total_factor = ifelse(kids_total==4, "4+", as.character(kids_total)),
         kids_total_factor = factor(kids_total_factor, levels = c("0", "1", "2", "3", "4+")))%>%
  mutate(scale_vol_HIPP= as.numeric(scale(vol_HIPP)),
         scale_vol_tiv = as.numeric(scale(vol_tiv)),
         vol_HIPP_cm = vol_HIPP/1000,
         vol_tiv_cm = vol_tiv/1000)%>%
  drop_na(Z_PACC_V2)%>%
  drop_na(scale_vol_HIPP)%>%
  drop_na(cog_diff)%>%
  mutate(Z_PACC_diff_rate = Z_PACC_diff/cog_diff,
         vol_HIPP_rate = vol_HIPP/vol_tiv)
  
```

```{r, eval = F, include=F}
library(mediation)


#all
test_med_all <- lm(Z_PACC_diff ~ A  + age_v1 + APOE_binary + education + cog_diff, 
                     data = data_all[data_all$parous_group=="2+",])
summary(test_med_all)


model_mediator_all <- lm(scale_vol_HIPP ~ A + age_v1 + APOE_binary + education + scale_vol_tiv, 
                     data = data_all)

# Model 2: Effect of parity and HV on cognition
model_outcome_all <- lm(Z_PACC_diff ~  scale_vol_HIPP + A + scale_vol_tiv+ cog_diff + age_v1 + APOE_binary + education, 
                    data = data_all)

# Check your models
summary(model_mediator_all)
summary(model_outcome_all)
set.seed(123)

mediation_results <- mediate(model_mediator_all,    # Mediator model
                              model_outcome_all,     # Outcome model
                              treat = "A",  # Independent variable
                              mediator = "scale_vol_HIPP",   # Mediator
                              boot = TRUE,       # Use bootstrapping
                              sims = 5000) 

summary(mediation_results)


#multiparous
test_med <- lm(Z_PACC_diff ~ A  + age_v1 + APOE_binary + education + cog_diff, 
                     data = data_all[data_all$parous_group=="2+",])
summary(test_med)


model_mediator <- lm(scale_vol_HIPP ~ A + age_v1 + APOE_binary + education + scale_vol_tiv, 
                     data = data_all[data_all$parous_group=="2+",])

# Model 2: Effect of parity and HV on cognition
model_outcome <- lm(Z_PACC_diff ~  scale_vol_HIPP + A + scale_vol_tiv+ cog_diff + age_v1 + APOE_binary + education, 
                    data = data_all[data_all$parous_group=="2+",])

# Check your models
summary(model_mediator)
summary(model_outcome)
set.seed(123)

mediation_results2 <- mediate(model_mediator,    # Mediator model
                              model_outcome,     # Outcome model
                              treat = "A",  # Independent variable
                              mediator = "scale_vol_HIPP",   # Mediator
                              boot = TRUE,       # Use bootstrapping
                              sims = 5000) 

summary(mediation_results2)



```

Mediation analyses testing whether HV mediates the association between Aβ status and mPACC based on parity, in the whole sample, showed no significant mediation effect of HV on the relationship between Aβ status and mPACC difference (p=0.226). The total effect was nearing significance (β=-0.111, CI\[-0.247,0.020\] , p=0.063), and the direct effect was not significant (p=0.09).

Looking at multiparous women only (n= `r nrow(data_all[data_all$parous_group=="2+",])`), mediation analyses showed that HV significantly mediated about 27% of the effect of Aβ on cognitive difference (Average Causal Mediation Effect (ACME) = -0.045, 95% CI \[-0.112, -0.001\], p=0.036) (Figure 3). The total effect was significant (β = -0.170, 95% CI \[-0.335, -0.003\], p=0.043), with approximately 27% of the effect mediated through HV. The direct effect was not significant (Average Direct Effect (ADE) = -0.125, 95% CI \[-0.290, 0.040\], p = 0.147).

# Discussion

In this study, we examined the associations between parity and cognitive performance and HV, and whether Aβ status modified these associations. Parity did not significantly predict Aβ status in our sample, but instead interacted with it to affect both cognitive changes and HV across timepoints. In Aβ-positive women, higher parity was linked with steeper cognitive decline but not in Aβ-negative women. Regarding HV, higher parity was associated with smaller volumes in Aβ-positive, but not in Aβ-negative, women. Sub-group analyses showed that main effects of the parity and Aβ status interaction were observed in multiparous women. Mediation analyses in the subgroup of multiparous women indicated that HV partially mediated the effect of Aβ status on cognitive change. Together, these findings highlight a potential long-term influence of previous pregnancies on AD-related processes in postmenopausal women, suggesting that parity could interact with Aβ pathology to shape cognitive outcomes, partly through brain structure.

An interesting result of our study was that parity did not independently affect cognitive changes, but instead, its effect was moderated by Aβ status. Our results suggest that the direction of parity’s effect on cognition may depend on the presence of AD pathological burden, and evoke a deleterious effect of higher parity on cognitive change only in the context of Aβ positivity. The effect size observed was small, but this is not surprising given the short average time period between visits and their clinical status at baseline. The direction of this interaction was present in both primiparous and multiparous women. Still, the comparison of predicted values between Aβ-negative and Aβ-positive groups was only significantly different in multiparous women, suggesting that pregnancy influences cognitive outcome in a cumulative way. This result offers an explanation for the mixed findings from previous studies examining the relationship between parity and cognitive decline , where sample composition and the lack of biomarker data might have obscured this interaction @Beeri2009a; @jung2020.

Similarly, parity did not predict HV independently, but it moderated the effect of amyloid load on HV. This indicates that the pathological burden conferred by amyloid accumulation on HV might be affected by previous reproductive history. Whilst HV is affected during pregnancy in first time mothers @martínez-garcía2021, the long-term consequences of this neuronal plasticity is unknown. Grand multiparity was found to be associated with lower HV in postmenopausal women but in a sample that included MCI women, who therefore might already have accumulated amyloid pathology @jung2020. In rats, parity prevented the age-related decline in hippocampal neural stem cells @duarte-guterman2023, so it is possible similar mechanisms could explain the direction of effect observed in Aβ-negative women. Interestingly, the effects observed in our analyses were independent of time, suggesting that the factors underlying these differences may have occurred before our window of observation, which excludes the menopausal transition. It is possible for some of this effect to be driven by the menopausal transition given the hippocampus' sensitivity to sex hormones and the importance of this transition for AD onset @brinton2015. Future research should further explore the menopausal transition and its relationship with HV.

Mediation analysis shows that, in multiparous women, about 27% of the effect of Aβ status on cognitive slopes is mediated through HV. This result suggests that, in multiparous women, Aβ may have a more adverse effect on cognitive functioning, and this effect might partly be mediated by HV loss. Aβ accumulation contributes to neurodegeneration in the hippocampus @cacciaglia2025a which is probably why we observe partial mediation through HV. But this mediation is only seen in multiparous women, suggesting a cumulative effect of AD pathophysiological features. Repeated periods of high plasticity during the gestational period or high cognitive demand during the child-rearing years could influence the terrain for neuropathology accumulation in later life. The mediation is only partial, meaning that other mechanisms are at play. Pregnancy has a broad impact that not only affects hippocampal structures but also grey matter volume and cortical thickness, which might also contribute to the results at hand @martínez-garcía2021. Therefore, future avenues of investigation should also examine whether the same interaction with Aβ load occurs in other major brain structures or is specific to the hippocampus.

Second, parity could interact with Aβ status through immune changes induced by pregnancy. Parity has long-lasting effects on the immune system, as evidenced by the presence of fetal cells in the mother’s tissue postpartum @Huang2021. A key theory behind sex differences in AD trajectory concerns immune differences and their effects on the inflammatory cascade that characterizes Aβ accumulation @buckley2022. Gestational-related immune changes could contribute to the observed sex differences in immunological pathways in AD, perhaps through microglial effects on Aβ accumulation. Third, pregnancy’s impact on the cardiovascular system could modify later-life cardiovascular risk, which is highly linked with all-cause dementia and AD risk @farnsworthvoncederwald2022. Pregnancy requires substantial vasculature changes for the well-being of the new foetus, in turn leading to significant increases in plasma volume and cardiac output @iacobaeus2017. Previous parity could alter participants' vascular system, which in turn could modulate the effect of Aβ on vasculature, resilience to Aβ accumulation, and AD risk and trajectory. Given that microglial activation is strongly associated with cerebrovascular injury and Aβ accumulation, it is possible that vascular and immune pathways interact.

Interestingly, our results highlight a group difference between primiparous and multiparous women, suggesting that biological and lifestyle factors operate differently across reproductive histories. Most human studies investigating the effects of pregnancy on brain and hippocampal structure have focused on first-time mothers @Hoekzema2017; @martínez-garcía2021; @pritschet2024, leaving open the question of whether pregnancy-related changes accumulate with each childbirth or occur primarily during the first pregnancy. But beyond the biological processes associated with pregnancy, it is also essential to consider the transition to motherhood itself, often referred to as matrescence, which involves substantial psychological, social, and lifestyle adjustments @servin-barthet2023a. These experiences, encompassing caregiving demands, stress exposure, and shifts in social roles and routines, may influence brain health through mechanisms distinct from those of pregnancy-related biology. Participants in the ALFA+ cohort mainly come from a homogeneous social background, and we found no significant differences in household income or occupation depending on the number of childbirths (eTable 5). Therefore, future research on the effect of pregnancy on women’s brain health should disentangle the biological versus social aspects of this phenomenon by using a more diverse population with higher variability of social backgrounds.

Some limitations should be considered. Firstly, the number of follow-up visits is insufficient to observe nonlinear evolution, thereby limiting our understanding of plausible mechanisms underlying AD pathogenesis. Future work should include additional timepoints to better understand the trajectory of these results. Secondly, although our sample’s excellent health record makes it ideal to identify subtle changes in the early AD continuum, the vast majority of ALFA+ study participants are white Caucasians, resulting in an ethnically homogeneous sample primarily from middle- to upper-middle-class backgrounds. Future work should reproduce these analyses in a more diverse cohort to understand the impact of ethnicity and social background on the present results. Thirdly, our sample had significantly more multiparous participants, meaning our sensitivity analyses using parity as a categorical variable were relatively underpowered for the nulliparous and primiparous levels. Therefore, Type II errors were more likely for these groups. Replication in larger independent cohorts is needed to address the relatively few cases of nulliparous and primiparous individuals. Additionally, the ALFA+ cohort is highly enriched in AD genetic phenotype @vilor-tejedor2023, with for example, nearly 40% of the participants being carriers of APOE-e4. This is not representative of the general population's genetic profile, and could have influenced our results due to carriers’ elevated vulnerability to AD pathology accumulation. Lastly, the absence of direct hormonal measurements limits our ability to draw definitive conclusions about the role of sex steroid hormones in the observed associations. By relying on proxy measures, we can only estimate hormone-related effects, potentially leading to an underestimation of their actual influence.

To conclude, this research demonstrated that parity is related to postmenopausal women's cognitive trajectory and hippocampal volume as a function of their Aβ status. While parity did not directly influence cognitive performance, it interacted with Aβ status to shape both cognitive changes and HV. By highlighting how reproductive history interacts with Aβ pathology, our results may help reconcile previously inconsistent findings on the association between parity and AD risk and underscore the importance of considering female-specific biological factors in AD research. In summary, this work contributes to a more nuanced understanding of why women exhibit more deleterious downstream effects of AD pathology and highlights the importance of incorporating sex and gender specific risk factors in prevention and early AD detection strategies.

# Appendix

## Regression tables


```{r}


tbl_regression(fit_cog_ashs, label= list(kids_total = "Parity (continous)", age_v1= "Age", APOE_binary = "APOE-E4", education = "Education (years of)", diff_mri = "Time", vol_HIPP_cm = "HV"), exponentiate = F,
               include=c(vol_HIPP_cm, diff_mri, kids_total, `vol_HIPP_cm:diff_mri`),
               estimate_fun = ~ style_number(.x, digits = 4))%>%
  modify_header(estimate ~ "**Estimate**", label ~ "**Variables, N = {N}**")
```


```{r parity_amyloid, echo=F, warning=F}

tbl_regression(fit_amyloid, label= list(kids_total = "Parity (continous)", age_v1= "Age", APOE_binary = "APOE-E4", education = "Education (years of)"), exponentiate = F,
               include=c(kids_total))%>%
  modify_header(estimate ~ "**Estimate**", label ~ "**Variables, N = {N}**")
```

```{r PACC_long_simple, echo=F, warning=F}

tbl_regression(fit_cog_PACC, 
               label= list(kids_total = "Parity (continous)", age_v1= "Age", APOE_binary = "APOE-E4", education = "Education (years of)", A= "Aβ+", time_diff_years="Time"), 
               show_single_row = c( A, `kids_total:time_diff_years`, `kids_total:A`, `A:time_diff_years`, `kids_total:A:time_diff_years`), 
               exponentiate = F, 
               include=c(kids_total, A, time_diff_years, `kids_total:A`, `A:time_diff_years`, `kids_total:time_diff_years`, `kids_total:A:time_diff_years`),
               estimate_fun = ~ style_number(.x, digits = 4))%>%
  modify_header(estimate ~ "**Estimate**", label ~ "**Variables, N = {N}**")


```

```{r PACC_long_int_cat, echo=F, warning=F}
tbl_regression(fit_cog_PACC_group, 
               label= list(parous_group = "Parity (categorical)", age_v1= "Age", APOE_binary = "APOE-E4", education = "Education (years of)", A= "Aβ+", time_diff_years="Time"), 
               show_single_row = c( A, `A:time_diff_years`), 
               exponentiate = F, 
               include=c(parous_group, A, time_diff_years, `A:time_diff_years`, `parous_group:A`, `parous_group:time_diff_years`, `parous_group:A:time_diff_years`),
               estimate_fun = ~ style_number(.x, digits = 4))%>%
  modify_header(estimate ~ "**Estimate**", label ~ "**Variables, N = {N}**")

```

```{r hippo_long, echo=F, warning=F}
tbl_regression(mod_ashs, label= list(kids_total = "Parity (continous)", age_v1= "Age", APOE_binary = "APOE-E4", education = "Education (years of)", A= "Aβ+", diff_mri="Time"), show_single_row = c(A, `kids_total:diff_mri`, `kids_total:A`, `diff_mri:A`, `kids_total:diff_mri:A`), exponentiate = F, include= c(kids_total, A, diff_mri, `kids_total:diff_mri`, `kids_total:A`, `diff_mri:A`, `kids_total:diff_mri:A`))%>%
  modify_header(estimate ~ "**Estimate**", label ~ "**Variables, N = {N}**")
```

```{r hippo_long_group, echo=F, warning=F}
tbl_regression(mod_ashs_group, label= list(parous_group = "Parity (categorical)", age_v1= "Age", APOE_binary = "APOE-E4", education = "Education (years of)", A= "Aβ+", diff_mri="Time"), show_single_row = c(A, `diff_mri:A`), exponentiate = F, include= c(parous_group, A, diff_mri, `parous_group:diff_mri`, `parous_group:A`, `diff_mri:A`, `parous_group:diff_mri:A`))%>%
  modify_header(estimate ~ "**Estimate**", label ~ "**Variables, N = {N}**")

```

## MRI data acquisition

MRI scans were acquired on a 3.0 T scanner (GE Discovery MR750 W 3 T) using the same protocol for all participants, which included one T1- and three T2- weighted sequences. The 3D-T1w sequence was designed with an isotropic voxel size of1mm3 and a matrix size of256x256x160 (TR/TE/TI = 8.0/3.7/450 ms, NSA = 1, flip angle = 8°). Three 3D-T2w sequences, with a voxel size of1 mm× 1 mm× 3 mm, were also used: fluid attenuation inversion recovery (FLAIR: TR/TE/ TI = 11,000/90/2600 ms, flip angle = 160°), fast spin echo (FSE: TR/TE = 5000/85 ms, flip angle = 110°) and, gradient echo (GRE: TR/TE = 1300/23 ms, flip angle = 15°). All scans were visually assessed to verify their quality and to detect incidental findings by a trained neuroradiologist and have been reported elsewhere @brugulat-serrat2017.

## Hippocampal subfields

```{r, include = F}
mod_ashs_CA1<-lm(vol_CA1 ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2 & data_ashs$Visit== 1,])
summary(mod_ashs_CA1)

mod_ashs_CA2<-lm(vol_CA2 ~ kids_total*A + age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_CA2)

mod_ashs_CA3<-lm(vol_CA3 ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_CA3)

mod_ashs_DG<-lm(vol_DG ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_DG)

mod_ashs_SUB<-lm(vol_SUB ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_SUB)

mod_ashs_ERC<-lm(vol_ERC ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_ERC)

mod_ashs_PHC<-lm(vol_PHC ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_PHC)

mod_ashs_BA35<-lm(vol_BA35 ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_BA35)

mod_ashs_BA36<-lm(vol_BA36 ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_BA36)

mod_ashs_sulc<-lm(vol_sulcus ~  kids_total*A +age_v1 +APOE_binary + scale(vol_tiv) + education , data = data_ashs[data_ashs$sex==2& data_ashs$Visit== 1,])
summary(mod_ashs_sulc)



```

No significant main effect of parity was found in any of the hippocampal subfields. However, there was a significant effect of the interaction between parity and Aβ on the dentate gyrus (β = `r tidy(mod_ashs_DG)$estimate[8]`, SE = `r tidy(mod_ashs_DG)$std.error[8]`, p = `r tidy(mod_ashs_DG)$p.value[8]`), and a trending effect of this same interaction in the subiculum (β = `r tidy(mod_ashs_SUB)$estimate[8]`, SE = `r tidy(mod_ashs_SUB)$std.error[8]`, p = `r tidy(mod_ashs_SUB)$p.value[8]`). The direction of the effect followed the same as for the whole hippocampus, where Aβ- women showed a positive relationship between parity and hippocampal volume, whilst Aβ- women showed the opposite (Figure X)

```{r, echo=F, warning=F}
preds_DG <- ggpredict(mod_ashs_DG, terms = c("kids_total", "A"))%>%
  as.data.frame()%>%
  mutate(group = ifelse(group==0, "Aβ-", "Aβ+"))

preds_SUB <- ggpredict(mod_ashs_SUB, terms = c("kids_total", "A"))%>%
  as.data.frame()%>%
  mutate(A = ifelse(group==0, "Aβ-", "Aβ+"))

DG<-preds_DG %>%
  ggplot(aes(x = x, y = predicted, color = group, group = group)) +
  geom_line(size=2)+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), colour=NA, alpha=0.2) +
  th +
  labs(x = "Number of childbirths", 
       y = "DG", 
       color = "Aβ Status") +
  theme(legend.position= "none")

SUB<-preds_SUB %>%
   ggplot(aes(x = x, y = predicted, color = group, group = group)) +
  geom_line(size=2)+
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), colour=NA, alpha=0.2) +
  th +
  labs(x = "Number of childbirths", 
       y = "SUB", 
       color = "Aβ Status") 


DG/SUB
```

## Chi square

```{r, echo=F, Warning=F}
library(MASS)

cont_tbl_ingreso<-table(data$kids_total,data$ingreso_hogar) 
cont_tbl_occupation<-table(data$kids_total,data$occupation) 

cont_tbl_ingreso
cont_tbl_occupation

summary(chisq.test(cont_tbl_ingreso))
tidy(chisq.test(cont_tbl_occupation))
```

# References

```{r, eval=F, include =F}
Third, both could interact by modifying cardiovascular risk. Cardiovascular risk is highly associated with all-cause dementia and AD risk @farnsworthvoncederwald2022, and pregnancy requires substantial vasculature changes for the well-being of the new foetus, in turn leading to significant increases in plasma volume and cardiac output @iacobaeus2017. To accommodate these changes, vascular resistance decreases through outward vessel hypertrophy. Although these changes are mostly seen in the uterine vasculature, these changes are, to some extent, pervasive to the entire vascular system @boeldt2017. Importantly, Aβ peptides can deposit on brain blood vessels, interfering with its integrity and function. Therefore, previous parity could modify participants' vascular system, which in turn could modulate the impact of Aβ on vasculature, or resilience to Aβ accumulation, and modify AD risk and trajectory. Given that microglial activation is strongly related to cerebrovascular injury, but also Aβ accumulation, it is possible that vascular and immune pathways interact with each other as well. Unfortunately, the effect of previous normotensive pregnancy on the vasculature system in older females is largely unknown, meaning further work is needed to characterise their vascular impact on postmenopausal individuals.

While it is widely agreed upon that the gestational period is associated with a sharp increase in sex steroids, notably E2 and P4, studies reporting this effect show staggering variability in measurement \[24\]. And an even greater uncertainty remains in the characterization of hormonal levels in the post-partum period and beyond. Some studies find that parous women show lower estrogen levels than nulliparous women \[26\] \[27\]. Still, this difference might only be evident 3 years after the last birth \[28\] and may depend on participants’ menstrual cycle at the time of measurement \[25\] and ethnicity \[29\]. Additionally, in rats, multiparity is associated with reduced sensitivity to estrogens compared with nulliparous rats \[30\]. Therefore, although previous parity might be associated with higher total lifetime exposure to estrogens, it might also be associated with lower circulating estrogen levels and lower sensitivity to it. Additionally, the menopausal transition adds another layer of complexity to interpretation, since it is unclear if the sharp decline in sex steroids correlates with premenopausal levels. Following this, although sex steroids certainly play a role in the interaction seen in our results, it is difficult to establish a direction of effect with certainty due to the remaining gaps in knowledge on the impact of parity on hormonal levels, especially in postmenopausal states.

Another theory is that parity interacts with Aβ status through immune changes induced by pregnancy. Parity affects the immune system beyond birth, as evidenced by the presence of fetal cells in the mother’s tissue postpartum \[31\]. The long-term consequences of these immune changes remain largely unknown, with current evidence limited to a few epidemiological studies examining the effect of parity on autoimmune disease \[32\]. A key theory behind sex differences in AD trajectory concerns immune differences and their effects on the inflammatory cascade that characterizes Aβ accumulation \[33\]. A postmortem study on 183 participants, of which two-thirds were clinically diagnosed with MCI or dementia before death, found that the percentage of Aβ accumulation mediated by microglial activation is nearly 3 times as high in females compared to males \[34\]. In that sense, gestational-related immune changes could contribute to the observed sex differences in immunological pathways in AD, perhaps through microglial effects on Aβ accumulation. Additionally, one study that included 93 postmenopausal women spanning the full range of possible CDR scores found that the cumulative number of first trimesters, but not third trimesters, was associated with reduced AD risk \[35\]. Since major immune changes during pregnancy happen during the first trimester, this further establishes a link between pregnancy-related immune changes and AD risk.
```




```{r, include=F, eval =F}
model_hv_A0 <- lm(vol_HIPP ~ age_v1 + education + vol_tiv + APOE_binary, 
                   data = data_all_A0_2)
model_cog_A0 <- lm(Z_PACC_diff ~ age_v1 + education + vol_tiv + APOE_binary, 
                    data = data_all_A0_2)

# Create dataframe with residuals
plot_data_A0 <- data.frame(
  HV_resid = residuals(model_hv_A0),
  Cog_resid = residuals(model_cog_A0),
  Group = "Aβ-"
)

model_hv_A1 <- lm(vol_HIPP ~ age_v1 + education + vol_tiv + APOE_binary, 
                   data = data_all_A1_2)
model_cog_A1 <- lm(Z_PACC_diff ~ age_v1 + education + vol_tiv + APOE_binary, 
                    data = data_all_A1_2)

# Create dataframe with residuals
plot_data_A1 <- data.frame(
  HV_resid = residuals(model_hv_A1),
  Cog_resid = residuals(model_cog_A1),
  Group = "Aβ+"
)

# Combine
plot_data_combined <- rbind(plot_data_A0, plot_data_A1)
```

```{r, include=F, eval=F}
# Plot
ggplot(plot_data_combined, aes(x = HV_resid, y = Cog_resid, color=Group)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, size=1.3) +
  facet_wrap(~Group, scale="free") +
  labs(
    x = "Hippocampal Volume (residuals)",
    y = "PACC Change (residuals)",
  ) +
  th+
  stat_cor(method = "pearson", 
           label.x.npc = "left", 
           label.y.npc = "top",
           size = 4,
           colour="black")+
  scale_color_manual(values = c("#1CA8B7", "#FF7043"))

```

```{r, include=F, eval=F}

preds_PACC <- ggpredict(fit_cog_PACC, terms = c("time_diff", "kids_total", "A"))

preds_PACC_A0<-as.data.frame(preds_PACC)%>%
  rename(kids_total_factor = group,
         A=facet)%>%
  filter(A==0)%>%
  mutate(A = ifelse(A==0, "Aβ-", "Aβ-"),
         kids_total_factor= ifelse(as.character(kids_total_factor)==4, "4+", as.character(kids_total_factor)))

preds_PACC_A1<-as.data.frame(preds_PACC)%>%
  rename(kids_total_factor = group,
         A=facet)%>%
  filter(A==1)%>%
  mutate(A = ifelse(A==1, "Aβ+", "Aβ+"),
         kids_total_factor= ifelse(as.character(kids_total_factor)==4, "4+", as.character(kids_total_factor)))

# Create a combined factor for color mapping
data_plot <- data_cog_long %>%
  filter(sex == 2, !is.na(A)) %>%
  mutate(kids_total_factor = ifelse(kids_total==4, "4+", kids_total),
         A = factor(A, levels = c("0", "1"), labels = c("Aβ-", "Aβ+")),
         color_group = paste(A, kids_total_factor, sep = "_"))

preds_PACC_A0 <- preds_PACC_A0 %>%
  mutate(color_group = paste("Aβ-", kids_total_factor, sep = "_"))

preds_PACC_A1 <- preds_PACC_A1 %>%
  mutate(color_group = paste("Aβ+", kids_total_factor, sep = "_"))


# Define color palettes
color_values <- c(
  "Aβ-_0" = "#4DD0E1", "Aβ-_1" = "#1CA8B7", "Aβ-_2" = "#1E8895", "Aβ-_3" = "#0D707B", "Aβ-_4+" = "#0F565E",
  "Aβ+_0" = "#FF8A65", "Aβ+_1" = "#FF7043", "Aβ+_2" = "#FF5722", "Aβ+_3" = "#F4511E", "Aβ+_4+" = "#BF360C"
)

ggplot(data_plot, aes(x = time_diff, y = Z_PACC, color = color_group)) +
  geom_point(alpha = 0.3, size=2) +
  geom_line(data = preds_PACC_A0, aes(x=x, y=predicted, group=kids_total_factor, color=color_group), size=1.3) +
  geom_line(data = preds_PACC_A1, aes(x=x, y=predicted, group=kids_total_factor, color=color_group), size=1.3) +
  geom_ribbon(data = preds_PACC_A0, aes(x=x, ymin=conf.low, ymax=conf.high, 
                                         group=kids_total_factor, fill=color_group), 
              inherit.aes = F, alpha=.15)+
  geom_ribbon(data = preds_PACC_A1, aes(x=x, ymin=conf.low, ymax=conf.high, 
                                         group=kids_total_factor, fill=color_group), 
              inherit.aes = F, alpha=.15)+
  facet_wrap(~A) +
  scale_color_manual(values = color_values, labels = c("0", "1", "2", "3", "4+", "0", "1", "2", "3", "4+")) +
  scale_fill_manual(values = color_values, labels = c("0", "1", "2", "3", "4+", "0", "1", "2", "3", "4+")) +
  labs(x = "Time difference (days)", 
       y = "PACC Z-score", 
       color = "Number of childbirths",
       fill = "Number of childbirths")+
  th+
  geom_text(data = data.frame(A = factor("Aβ+", 
                                                     levels = c("Aβ-", "Aβ+")), 
                              time_diff = Inf, 
                              Z_PACC = Inf, 
                              label = "p= 0.033"),
            aes(x = time_diff, y = Z_PACC, label = label),
            size = 4, vjust = 1.5, hjust = 1.2, inherit.aes = FALSE)
```

```{r, eval=F, include=F}
tbl_regression(mod_ashs_DG, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))

tbl_regression(mod_ashs_SUB, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))%>%
  modify_caption("**Subiculum**")


tbl_regression(mod_ashs_CA1, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))%>%
  modify_caption("**CA1**")

tbl_regression(mod_ashs_CA2, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))%>%
  modify_caption("**CA2**")

tbl_regression(mod_ashs_CA3, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))%>%
  modify_caption("**CA3**")

tbl_regression(mod_ashs_PHC, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))%>%
  modify_caption("**Para hippocampal coretx**")

tbl_regression(mod_ashs_ERC, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))%>%
  modify_caption("**Entorhinal cortex**")

tbl_regression(mod_ashs_BA35, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))%>%
  modify_caption("**Brodmann area 35**")

tbl_regression(mod_ashs_BA36, show_single_row = c(A, APOE_binary, `kids_total:A`), 
               label = list(A = "Aβ+", APOE_binary = "APOE4 Carrier", `scale(vol_tiv)`= "TIV", kids_total= "Number of childbirths"))%>%
  modify_caption("**Brodmann area 36**")
  
```

```{r, eval=F, include=F}
library(dagitty)
library(ggdag)

set.seed(123)

g <- dagitty("dag{  
  APOE_binary -> {A vol_Whole_hippocampus}
  age_v1 -> {kids_total vol_Whole_hippocampus A}
  education -> {kids_total vol_Whole_hippocampus A};
  A -> vol_Whole_hippocampus
  kids_total -> {Hormonal_exposure}
  APOE4 -> {A vol_Whole_hippocampus}
  Hormonal_exposure -> {vol_Whole_hippocampus APOE4}
  kids_total [exposure]
  vol_Whole_hippocampus [outcome]
  Hormonal_exposure [unobserved]
 }")

coords<-list(
  x=c(age_v1 = 1, APOE4 = 1, kids_total = 2, education = 2, Hormonal_exposure = 3, A= 3, vol_Whole_hippocampus = 4),
  y=c(age_v1 = -2, APOE4 = 2, kids_total = -1, education = 1, A = 1, Hormonal_exposure = -1,vol_Whole_hippocampus = 0))

dag_r<-dagify(vol_Whole_hippocampus ~ kids_total + education + age_v1 + A + APOE4 + age_v1 + Hormonal_exposure,
              A ~ age_v1 + APOE4 + CAIDE_scr_20M1,
              kids_total ~ education + age_v1,
              Hormonal_exposure ~ kids_total + age_v1,
              labels = c(kids_total = "Parity",
                         APOE4 = "APOE-E4 Carriership",
                         education = "Education",
                         A = "Aβ status",
                         vol_Whole_hippocampus = "Whole hippocampal volume",
                         age_v1 = "Age at baseline",
                         Hormonal_exposure = "Hormonal exposure"),
              exposure = "kids_total",
              outcome = "vol_Whole_hippocampus",
              coords = coords)

dag_r_tidy<-tidy_dagitty(dag_r, layout="nicely")%>%
  node_dconnected(controlling_for = c("age_v1", "education"))
```

```{r, eval=F, include=F}
dag_r_tidy%>%
  mutate(adjusted=str_to_title(adjusted),
         arrow=ifelse(adjusted == "Adjusted", 0.15, 0.85))%>%
  ggplot(aes(x=x, y=y, xend=xend, yend=yend, colour=adjusted, fill=adjusted, shape=adjusted))+
  geom_dag_point()+
  geom_dag_edges(aes(edge_alpha=arrow), edge_width=0.7)+
  geom_dag_label_repel(aes(label = label), colour = "white", show.legend = FALSE)+
  theme_dag() +
  theme(legend.position= "none")+
  scale_shape_manual(values = c(22, 21))
```

```{r, include=F, eval=F}
data_juando<-data_ashs%>%
  filter(sex==2, Visit==1)

followup<-data_ashs%>%
  filter(Visit==2, sex==2)%>%
  rename(followup = vol_HIPP)%>%
  dplyr::select(subject, followup)

data_juando<-left_join(data_juando, followup)

test<-lm(followup~  kids_total*A  +APOE_binary + education + scale(vol_tiv) , data = data_juando)

summary(test)
```


